---
title: "Lab 10 - Grading the professor, Pt. 2"
author: "Linn Zapffe"
date: "19/03/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

### Exercise 1

Fitting a linear model between professor rating and beauty score:
```{r reg-rating-beauty}
m_bty <- lm(score ~ bty_avg, data = evals)
m_bty
```

The linear model is teacher rating = 3.88 + 0.067 * beauty rating.

```{r sum-rating-beauty}
summary(m_bty)
```

The R^2 is .035 and the adjusted R^2 is .033. So 3.5 or 3.3% of the variance in professor rankings is explained by beauty scores.

### Exercise 2

Fitting a linear model between professor rating and beauty score + gender:
```{r reg-rating-beauty-gender}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
m_bty_gen
```
The model is professor rankings = 3.75 + 0.074 * beauty ranking + 0.17 (if the professor is male).

```{r sum-rating-beauty-gender}
summary(m_bty_gen)
```

The R^2 is .059 and the adjusted R^2 is .055. So 5.9 or 5.5% of the variance in professor rankings is explained by beauty scores and the professors' gender.


## Exercise 3

I didn't fit this model with an interaction, so I will fit another model with the interaction.

Fitting a linear model between professor rating and beauty score + gender, with an interaction:
```{r reg-rating-beauty-gender-int}
m_bty_gen_int <- lm(score ~ bty_avg + gender + bty_avg*gender, data = evals)
m_bty_gen_int
```

The slope is in this case the professor ranking if they have a beauty score of 0 and are female. The interaction is how much more of a boost in ranking the professor gets for each additional beauty score because they are male (so in addition to the 0.031 boost per beauty score).

## Exercise 4

```{r sum-rating-beauty-gender-int}
summary(m_bty_gen_int)
```

According to our R^2 value, 7.1% of the variance in professor rankings are explained by our model: beauty scores, gender, and the interaction between beauty scores and gender. If we use the adjusted R^2, it is 6.5%.


## Exercise 5

The professor rankings of male professors is:

professor rankings = 3.95 + 0.031 * beauty ranking - 0.18 (if the professor is male) + 0.080 per beauty ranking if professor is male.

The equation if the professor is male is 3.95 + 0.031 * beauty ranking - 0.18 + 0.080 per beauty ranking if professor is male
= 3.77 + 0.031 * beauty ranking + 0.080 per beauty ranking if professor is male


## Exercise 6

For professors with the same beauty score, the males scores higher, as they get a general penalty of -0.18 points, but also get 0.080 additional points for the ranking per beauty score. So unless the professor in question get ranked as very low in beauty, male professors will get a higher score than female professors with the same beauty rankings


## Exercise 7

There is an interaction effect of + 0.080 for men, which means that there is a stronger relationship between beauty scores and professor rankings for male than female professors.


## Exercise 8

The R^2 for m_bty_gen_int is 0.071, while it is .035 for m_bty. The model with beauty, gender and the interaction therefore seem to explain twice the amount of variance in the professor rankings compared to the model using just the beauty scores.

Gender (and its interaction with beauty scores) therefore seem to explain an additional 0.036 of the variance in professor rankings.


## Exercise 9

The slope of beauty in the m_bty_gen_int model is 0.031, while it is 0.067 in m_bty. The addition of gender decreased the general slope, but this would be if only considering female professors. If looking at male professors, you would have to also look at the interaction, which adds 0.080 to the slope, giving a steeper slope for males in the m_bty_gen_int model compared to m_bty.


## Exerxise 10

Fitting a linear model between professor rating and beauty score + rank, with an interaction:
```{r reg-rating-beauty-rank-int}
m_bty_rank_int <- lm(score ~ bty_avg + rank + bty_avg*rank, data = evals)
m_bty_rank_int
```

The equation is professor ranking = 4.10 + 0.04 per beauty point - 0.02 if on tenure track - 0.41 if tenured - 0.03 if tenure track per beauty point + 0.07 if tenured per beauty point.

The intercept is the professor ranking of a teaching professor that has a beauty score of 0, which is 4.10.

The slope is how much better the rating get per beauty score for teaching professors, which is 0.04.

If the professor is tenure track, they get a penalty of -0.02 to the intercept and -0.03 to the slope (penalty per beauty score).

If the professor is tenured, they get a penalty of -0.41 to the intercept and -0.07 to the slope (penalty per beauty score).


## Exercise 11

I think the worst predictor would be cls_profs, as I don't think how many professors are teaching the sections would have that much of an impact on professor rankings. It might have some predictive power, but I don't think it would be much.


## Exercise 12

Fitting a linear model between professor rating and number of professors teaching sections:
```{r reg-rating-beauty-num-prof}
m_bty_num_prof <- lm(score ~ bty_avg + cls_profs, data = evals)
m_bty_num_prof
```

If the professor is the only one teaching sections in a course, the professor ranking decreases by 0.04 points. This is not nothing, but not a lot. There could be other predcitors that explain less variance in the data set though.


## Exercise 13

If I am including cls_perc_eval and cls_students in my model, I would not want to include cls_did_eval. This is because cls_perc_eval is the percent of students completing the evaluation, whicl cls_students is the number of students in the class. The product of these two variables would be equal to the variable cls_did_eval. We therefore do not want to include this variable as it would be perfectly related to the other two variables and the predictive power of them would be masked as the estimates only show unique predictive power.


## Exercise 14

Fitting a linear model between professor rating and all variables except for cls_did_eval:
```{r reg-rating-all}
m_bty_all <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval + cls_students + cls_level + cls_profs + cls_credits + bty_avg, data = evals)
m_bty_all
```

Currently, it seems like the number of students in the class has the lowest predictive power, with 0.0004 points to the rankings. This is a little surprising, so before making this conclusion as a definite, I would look into the relationship of this variables with others, such as cls_level, as there could be some relationship between the cls_students variable and other variable in the data set that decreases the estimate for the cls_students predictor in the model.


## Exercise 15

Fitting a linear model between professor rating and all variables except for cls_did_eval, then going backwards and removing variables:
```{r reg-rating-all-backwards-estimate}
m_bty_all_backwards <- lm(score ~ ethnicity + gender + language + cls_credits + bty_avg, data = evals)
summary(m_bty_all_backwards)
```

Fitting a linear model between professor rating and all variables except for cls_did_eval, then going backwards and removing variables::
```{r reg-rating-all-backwards-p}
m_bty_all_backwards <- lm(score ~ ethnicity + gender + cls_perc_eval + cls_credits + bty_avg, data = evals)
summary(m_bty_all_backwards)
```

I tried to approaches since my first one didn't work that well. Originally, I removed the predictors with the lowest estimates one by one. Then next time, I removed then based on p-values, which worked better. The second model explain 13.66% of the variance, while the first explains 11.70% of the variance in professor rankings.

Professor rankings = 3.14 + 0.23 if professor is not an ethnic minority + 0.16 if professor is male + 0.005 per percent of students completed the evaulation 0.54 if class is one credit + 0.74 per beauty score.


## Exercise 16

The slope for not being a minority is 0.233794, so if the professor is not an ethnic minority, they get a boost in ranking of 0.23, while if they are a minority, they do not get this boost.

The slope for bty_avg is 0.073, so for each additional beauty point, the ranking increases by 0.073.


## Exercise 17

According to the final model, a professor would get a good ranking at the University of Austin Texas if they are of not an ethnic minority, male, and attractive, as well as if the class is one credit and has a high percentage of students filling out the evaluations.


## Exercise 18

There is probably some truth to these characteristics and the rankings professors get. However, there might also be some characteristics that are more specific to the University of Austin Texas due to its location and characteristics of the student body. There would maybe for example not be as much of a difference in etnic minority status in an university that has a larger percent of their student body that is an ethinc minority themselves. I would therefore not generalize it to all other universities, but rather instead consider it a general suggestion of what seems to matter for professor rankings, which require more broad data or data specific to the university of interest to draw more certain conclusions.


