---
title: "Lab 10 - Grading the professor, Pt. 2"
author: "Linn Zapffe"
date: "19/03/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

### Exercise 1

Fitting a linear model between professor rating and beauty score:
```{r reg-rating-beauty}
m_bty <- lm(score ~ bty_avg, data = evals)
m_bty
```

The linear model is teacher rating = 3.88 + 0.067 * beauty rating.

```{r sum-rating-beauty}
summary(m_bty)
```

The R^2 is .035 and the adjusted R^2 is .033. So 3.5 or 3.3% of the variance in professor rankings is explained by beauty scores.

### Exercise 2

Fitting a linear model between professor rating and beauty score + gender:
```{r reg-rating-beauty-gender}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
m_bty_gen
```
The model is professor rankings = 3.75 + 0.074 * beauty ranking + 0.17 (if the professor is male).

```{r sum-rating-beauty-gender}
summary(m_bty_gen)
```

The R^2 is .059 and the adjusted R^2 is .055. So 5.9 or 5.5% of the variance in professor rankings is explained by beauty scores and the professors' gender.


## Exercise 3

I didn't fit this model with an interaction, so I will fit another model with the interaction.

Fitting a linear model between professor rating and beauty score + gender, with an interaction:
```{r reg-rating-beauty-gender-int}
m_bty_gen_int <- lm(score ~ bty_avg + gender + bty_avg*gender, data = evals)
m_bty_gen_int
```

The slope is in this case the professor ranking if they have a beauty score of 0 and are female. The interaction is how much more of a boost in ranking the professor gets for each additional beauty score because they are male (so in addition to the 0.031 boost per beauty score).

## Exercise 4

```{r sum-rating-beauty-gender-int}
summary(m_bty_gen_int)
```

According to our R^2 value, 7.1% of the variance in professor rankings are explained by our model: beauty scores, gender, and the interaction between beauty scores and gender. If we use the adjusted R^2, it is 6.5%.


## Exercise 5

The professor rankings of male professors is:

professor rankings = 3.95 + 0.031 * beauty ranking - 0.18 (if the professor is male) + 0.080 per beauty ranking if professor is male.

THIS QUESTION IS NOT FINISHED