---
title: "Lab 02 - Plastic waste"
author: "Linn Zapffe"
date: "24/01/2025"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

Remove this text, and add your answer for Exercise 1 here.

```{r plastic-waste-continent}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap)) +
  geom_histogram() +
  facet_wrap(~ continent)
```

```{r demon 1}
ggplot(
  data = plastic_waste,
  aes(x = plastic_waste_per_cap)
) +
  geom_density()
```

```{r demon 2}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent
  )
) +
  geom_density()
```

```{r demon 3}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density()
```

```{r demon 4}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density(alpha = 0.7)
```


### Exercise 2

```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density(alpha = 0.45)
```

What we add to aes() is what variables we somehow want to divide up our data by. So there is a difference for each instance. For example, an axis with all possible scores or a color for each continent. The things we add in the geom, such as alpha, is the same for all the data.

### Exercise 3

```{r demon 5}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_boxplot()
```

Changing it to violin plots

```{r plastic-waste-violin}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_violin()
```

It is a lot easier to see the distributions with the violing plots than the box plots. With the box plots you have to interpret where the median is relative to the interquartiles, while for the violin plots, it is very easy to see whether there is more data for lower or higher values just by looking at the shape of it. You also see "all" the scores in the distribution and not just the 5 summary statistics.


### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r plastic-waste-mismanaged}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap)) + 
  geom_point()
```

There doesn't seem to be that much of a relationship between plastic waste per capita and mismanaged plastic waste capita. There is a weak positive relationship. The relationship seems fairly consistent though.

```{r plastic-waste-mismanaged-continent}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap, color = continent)) + 
  geom_point()
```

I don't see any clear distinctions for this relationship based on continent. Except for the fact that Africa has a very clear (but weak) positive relationship. The other continents seem to have a relationship that is a little less consistent.

```{r plastic-waste-population-total}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = total_pop)) + 
  geom_point()
```

```{r plastic-waste-population-coastal}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = coastal_pop)) + 
  geom_point()
```

The relationship for both of these graphs is still fairly weak. However, there seems to be more of a relationship for coastal population, which has more of a line, than total population, where most of the data is centered for lower values of the total population and some centered at higher values, but fewer in the middle of the relationship.

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
# insert code here
```


