---
title: "Lab 08 - University of Edinburgh Art Collection"
author: "Linn Zapffe"
date: "21/02/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
# Library for plots, data handling, and web scraping
library(tidyverse) 
library(skimr)
# Library to check access to the web page URL through web scraping
library(robotstxt)
```

Checking that we have access to web scrape from the web page with the data:
```{r path-permission}
paths_allowed("https://collections.ed.ac.uk/art)")
```

As seen, we have access to web scrape from the web page, so we can move forward with out project.




```{r load-data, message = FALSE, eval = FALSE}
# Remove eval = FALSE or set it to TRUE once data is ready to be loaded
uoe_art <- read_csv("data/uoe-art.csv")
```

### Exercise 9

```{r separate-title-date, error = TRUE}
uoe_art <- uoe_art %>%
  separate(title, into = c("title", "date"), sep = "\\(") %>%
  mutate(year = str_remove(date, "\\)") %>% as.numeric()) %>%
  select(title, artist, year, ___)
```

### Exercise 10

Remove this text, and add your answer for Exercise 10 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 11

...

Add exercise headings as needed.
