---
title: "Lab 01 - Hello R"
author: "Linn Zapffe"
date: "16-01-2025"
output: 
  html_document: 
    fig_height: 7
    highlight: tango
    theme: flatly
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(datasauRus)
```

## Exercises

### Exercise 1

The datasaurus_dozen data set has 1 846 rows and 3 variables (columns)
The three variables are the data set (out of the 13 options),  the x values, and
the y values.

### Exercise 2

First, let's plot the data in the dino dataset:

```{r plot-dino, fig.height=6, fig.width=6}
# Filtering the overall data set to only contain the data set "dino" and save this in a variable called "dino_data"
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")

# Plotting the "dino" data
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

The plot shows a dinosaur.

Next, let's calculate the correlation between `x` and `y` in this dataset:

```{r cor-dino}
dino_data %>%
  summarize(r = cor(x, y))
```
The correlation of the dinosaur data set is -0.064, so close to 0.

### Exercise 3

Plotting the star data, using the code for the dino data as a template.

```{r plot-star}
# Filtering the overall data set to only contain the data set "star" and save this in a variable called "star_data"
star_data <- datasaurus_dozen %>%
  filter(dataset == "star")

# Plot the "star" data
ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

The star data looks like a star plotted.


```{r cor-star}
star_data %>%
  summarize(r = cor(x, y))
```
The correlation for the star data is -0.063, which is very similar to that of the dinosaur data.


### Exercise 4

Add your code and narrative to complete this exercise. Use the conventions from the previous exercises to name your chunks and organize your work.

```{r plot circle}
# Filtering the overall data set to only contain the data set "circle" and save this in a variable called "circle_data"
circle_data <- datasaurus_dozen %>%
  filter(dataset == "circle")

# Plot the "circle" data
ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

The circle data does, not surprisingly, look like a circle.

```{r circle cor}
circle_data %>%
  summarize(r = cor(x, y))
```

The correlation for the circle data is -0.068, which is also very similar to that 
of the dinosaur and the star.

### Exercise 5

First, we plot all the data sets, separating them by color.

```{r plot all data sets}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset)) +
  geom_point() +
  facet_wrap(~dataset, ncol = 3) +
  theme(legend.position = "none")
```

Then, we calculate the correlation for all of the data sets.

```{r correlation all data sets}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y)) %>%
  # Took out the 13 inside of print because it made it print only the data set names when knitting the document. Now, I get the correlations.
  print()
```

All the correlations are somewhere around -0.06.


Played around and changed the syntax highlighting to tango and the theme to flatly.
