---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Linn Zapffe"
date: "5/2/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r filter_d_Alaska}
dennys_ak <- dennys %>%
  filter(state == "AK")
nrow(dennys_ak)
```

There are `r nrow(dennys_ak)` locations of Denny's in Alaska.

```{r filter_l_Alaska}
laquinta_ak <- laquinta %>%
  filter(state == "AK")
nrow(laquinta_ak)
```

There are `r nrow(laquinta_ak)` locations of La Quinta in Alaska.

### Exercise 2

```{r comb_calc}
comb <- nrow(laquinta_ak) * nrow(dennys_ak)
```
There are `r comb` combinations of Denny's and La Quintas.

### Exercise 3

```{r join_dfs}
dn_lq_ak <- full_join(dennys_ak, laquinta_ak, by = "state")
dn_lq_ak
```


### Exercise 4

There are `r nrow(dn_lq_ak)` observations in the combined data frame.
The variable names are `r names(dn_lq_ak)`.

### Exercise 5

The following function is copy-pasted from the haversine.R file somewhere on the internet.
```{r define_haversine}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}
```


### Exercise 6

Calculate the distance between the locations:

```{r calc_dist}
dn_lq_ak <- dn_lq_ak %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y))
```



### Exercise 7

```{r min_dist}
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
dn_lq_ak_mindist
```


### Exercise 8

```{r vis_ak}
dn_lq_ak_mindist %>%
  ggplot(aes(x = address.x, y = closest, fill = address.x)) +
  # Had to include stat = "identity" so that it doesn't try to calculate counts, since we only have one data point per address
  geom_bar(stat = "identity") +
  labs(
    title = "Smallest Distances Between La Quintas and Denny's in Alaska",
    x = "Address of Denny's",
    y = "Distance to closest La Quinta (km)"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  theme_minimal()
```

The distances between the Denny's and La Quintas in Alaksa seem to be not too long, as the average distance is `r mean(dn_lq_ak_mindist$closest)`km. It would therefore take on average maybe 5 minutes to drive between the La Quintas and Denny's. However, it is not walking distance.


### Exercise 9

Doing this all again, but for North Carolina instead of Alaska:

Filter for locations in North Carolina and join the data frames for the two locations:
```{r comb_df_nc}
dennys_nc <- dennys %>%
  filter(state == "NC")
nrow(dennys_nc)

laquinta_nc <- laquinta %>%
  filter(state == "NC")
nrow(laquinta_nc)

dn_lq_nc <- full_join(dennys_nc, laquinta_nc, by = "state")
dn_lq_nc
```

Calculating distances between the locations:
```{r calc_dist_nc}
dn_lq_nc <- dn_lq_nc %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y))

dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

Visualizing the smalles distances between Denny's and La Quintas:
```{r vis_nc}
dn_lq_nc_mindist %>%
  ggplot(aes(x = closest)) +
  geom_histogram(fill = "darkblue") +
  labs(
   title = "Smallest Distances Between La Quintas and Denny's in North Carolina",
   x = "Minimum distance between Denny's and La Quintas (km)",
   y = "Number of locations with similar distance") +
  theme_minimal()
```

The distances are a lot bigger between the two chains in North Carolina compared to Alaska. The average distance is `r mean(dn_lq_nc_mindist$closest)` km away from each other, which is a solid drive, maybe around to 1-2 hours from each other.


### Exercise 10

Doing this all again, but this time, for Texas:

Filtering for locations in Texas and joining the two data frames:
```{r comb_df_tx}
dennys_tx <- dennys %>%
  filter(state == "TX")
nrow(dennys_tx)

laquinta_tx <- laquinta %>%
  filter(state == "TX")
nrow(laquinta_tx)

dn_lq_tx <- full_join(dennys_tx, laquinta_tx, by = "state")
dn_lq_tx
```

Calculate distances between the two locations:
```{r calc_dist_tx}
dn_lq_tx <- dn_lq_tx %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y))

dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

Visualizing the closest distances between Denny's and La Quintas:
```{r vis_tx}
dn_lq_tx_mindist %>%
  ggplot(aes(x = closest)) +
  geom_histogram(fill = "darkblue") +
  labs(
   title = "Smallest Distances Between La Quintas and Denny's in Texas",
   x = "Minimum distance between Denny's and La Quintas (km)",
   y = "Number of locations with similar distance") +
  theme_minimal()
```

The distances in Texas are a lot smaller than what it was in North Carolina. There is a little bit of range though, so some locations have the chains further apart from each other. However, the mean is `r mean(dn_lq_tx_mindist$closest)` km away from each other, which is similar to the average in Alaska, with about a 5 minute drive.


### Exercise 11

Doing this all again, this time for New York, where I went for undergrad:

Filtering by New York and joining the two data frames:
```{r comb_df_ny}
dennys_ny <- dennys %>%
  filter(state == "NY")
nrow(dennys_ny)

laquinta_ny <- laquinta %>%
  filter(state == "NY")
nrow(laquinta_ny)

dn_lq_ny <- full_join(dennys_ny, laquinta_ny, by = "state")
dn_lq_ny
```

Calculating distances and minimum distance between the two chains:
```{r calc_dist_ny}
dn_lq_ny <- dn_lq_ny %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y))

dn_lq_ny_mindist <- dn_lq_ny %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

Visualizing the minimum distance between the two chains:
```{r vis_ny}
dn_lq_ny_mindist %>%
  ggplot(aes(x = closest)) +
  geom_histogram(fill = "darkblue") +
  labs(
   title = "Smallest Distances Between La Quintas and Denny's in New York",
   x = "Minimum distance between Denny's and La Quintas (km)",
   y = "Number of locations with similar distance") +
  theme_minimal()
```

The distances in New York are more spread out than in Texas, but less so than in North Carolina. There are some that are very close to each other, some that are about an 1 hour away, and some that are even further away. The average is `r mean(dn_lq_ny_mindist$closest)` km, which is about half an hour of a drive.


### Exercise 12

Based on these analyses, it seems like whether Mitch Hedberg's joke is accurate or not depends on what state you are in. For some states, like Alaska and Texas, it seems pretty accurate, but for others, like North Carolina, it seems to not be very accurate.

