---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Linn Zapffe"
date: "5/2/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData)

# To highlight certain lines in a graph
library(ggrepel)
```

## Instructional staff employment trends

### Exercise 1

```{r load_staff_df}
staff <- read_csv("data/instructional-staff.csv")
```

```{r povit_staff}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

Line graph of the data set:

```{r standard_line_plot}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 2

Improving the above plot:

I tried first doing this without additional libraries. After getting labels that overlapped, I gave up and used the library ggrepel. That seems easier than to start adjusting labels individually.
```{r staff_line_graph}
staff_long %>%
  mutate(
    color_crit = case_when(
      faculty_type == "Part-Time Faculty" ~ "red",
      faculty_type %in% c("Full-Time Tenured Faculty", "Full-Time Tenure-Track Faculty", "Full-Time Non-Tenure-Track Faculty") ~ "darkblue",
      faculty_type == "Graduate Student Employees" ~ "grey"
    )
  ) %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = color_crit
  )) +
  geom_line(linewidth = 1) +
  scale_color_identity() +
  theme_minimal() +
  labs(
    title = "Percentage of Hires per Faculty Type over Time",
    x = "Year",
    y = "Proportion of hires",
    color = "Faculty type"
  ) +
  geom_label_repel(aes(label = if_else(year == 2003, faculty_type, "")))
```

I am trying to highlight that there are more part-time faculty, and fewer full-time faculty. Therefore, I chose to color the part-time faculty as being red, to draw the viewer's attention there. I colored all the full-time positions as blue, because the distinction doesn't matter that much in this case and left graduate student employees grey as that is not as important. I could also have collapsed the data for all full-time faculty and removed that for part-time faculty.

### Exercise 3

Here are some thoughts about how to improve the Fisheries plot:

- Not using 3D for the plots, because it is distracting and harder to interpret (area vs surface)
- Not use pie charts, but instead bar charts, because we are better at interpreting surface than area
- Maybe change the scale of the y-axis of the filled-line plot, since I can't see anything for most of the countries, since China is an extreme outlier. Alternatively, zoom in on the data with smaller values in China isn't that important.
- Remove the background color for the pie charts, makes it harder to see contrast
- Add a legend to the filled-line graph. I don't know what the two colors mean
- Add titles to the graphs
- Add a caption with the source of the data
- Remove shading from the pie chart, as that is only distracting
- Maybe add some texture in addition to the colors to make it more accessible and easier to read in grey scale printing. Alternatively, make colors that contrast more or add labels where the data is instead of having a legend.
- It is a little hard to know exactly what the graphs show, since it isn't explained well. However, it seems like the filled-line graph and pie chart is showing the same information, which seems unnecessary. I might make both, but I would think only one of them would be needed.
- I would also test out flipping the axes for the filled-line graph, since it is a little hard to read the country names when they are tilted.
- Not have more colored area that countries. The pie charts have 6 countries, but maybe 30 different sections. Are the country data divided into several categories? In that case, it should be collapsed into just one data point per country.

```{r load_fisheries_data}
fisheries <- read_csv("data/fisheries.csv")
```

Looking at the data, there are way to many countries to graph. Therefore, I will try seeing what the top countries are based on total fishery production:

```{r arrange_top}
fisheries %>%
  arrange(desc(total))
```

I think top 10 countries might be good. There might be political reasons to include other specific countries, but I don't have any of those, so I will stick with top 10. 

```{r filter_top_10}
fisheries_top_10 <- fisheries %>%
  slice_max(order_by = total, n = 10)
fisheries_top_10
```

Making a bar graph:
```{r plot_fish_bar_graph}
fisheries_top_10 %>%
  mutate(total = total / 100000) %>%
  ggplot(aes(x = reorder(country, -total), y = total, fill = country)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    total = "Total Fish Harvest by Top 10 Countries",
    x = "Countries",
    y = "Total Harvest (in 100 000s)"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

It could also be interesting to see the distribution of fish captures and harvested for each country in the bar graph. I will make that version of the plot too. 

```{r dist_fish_bar_graph}
# You need to make it long from wide by using pivot before trying to make the stacked bar graph.
```


### Exercise 4

...

Add exercise headings as needed.
