---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Linn Zapffe"
date: "5/2/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData)

# To highlight certain lines in a graph
library(ggrepel)
```

## Instructional staff employment trends

### Exercise 1

```{r load_staff_df}
staff <- read_csv("data/instructional-staff.csv")
```

```{r povit_staff}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

Line graph of the data set:

```{r standard_line_plot}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 2

Improving the above plot:

I tried first doing this without additional libraries. After getting labels that overlapped, I gave up and used the library ggrepel. That seems easier than to start adjusting labels individually.
```{r staff_line_graph}
staff_long %>%
  mutate(
    color_crit = case_when(
      faculty_type == "Part-Time Faculty" ~ "red",
      faculty_type %in% c("Full-Time Tenured Faculty", "Full-Time Tenure-Track Faculty", "Full-Time Non-Tenure-Track Faculty") ~ "darkblue",
      faculty_type == "Graduate Student Employees" ~ "grey"
    )
  ) %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = color_crit
  )) +
  geom_line(linewidth = 1) +
  scale_color_identity() +
  theme_minimal() +
  labs(
    title = "Percentage of Hires per Faculty Type over Time",
    x = "Year",
    y = "Proportion of hires",
    color = "Faculty type"
  ) +
  geom_label_repel(aes(label = if_else(year == 2003, faculty_type, "")))
```

I am trying to highlight that there are more part-time faculty, and fewer full-time faculty. Therefore, I chose to color the part-time faculty as being red, to draw the viewer's attention there. I colored all the full-time positions as blue, because the distinction doesn't matter that much in this case and left graduate student employees grey as that is not as important. I could also have collapsed the data for all full-time faculty and removed that for part-time faculty.

### Exercise 3

Here are some thoughts about how to improve the Fisheries plot:

- Not using 3D for the plots, because it is distracting and harder to interpret (area vs surface)
- Not use pie charts, but instead bar charts, because we are better at interpreting surface than area
- Maybe change the scale of the y-axis of the filled-line plot, since I can't see anything for most of the countries, since China is an extreme outlier. Alternatively, zoom in on the data with smaller values in China isn't that important.
- Remove the background color for the pie charts, makes it harder to see contrast
- Add a legend to the filled-line graph. I don't know what the two colors mean
- Add titles to the graphs
- Add a caption with the source of the data
- Remove shading from the pie chart, as that is only distracting
- Maybe add some texture in addition to the colors to make it more accessible and easier to read in grey scale printing. Alternatively, make colors that contrast more or add labels where the data is instead of having a legend.
- It is a little hard to know exactly what the graphs show, since it isn't explained well. However, it seems like the filled-line graph and pie chart is showing the same information, which seems unnecessary. I might make both, but I would think only one of them would be needed.
- I would also test out flipping the axes for the filled-line graph, since it is a little hard to read the country names when they are tilted.
- Not have more colored area that countries. The pie charts have 6 countries, but maybe 30 different sections. Are the country data divided into several categories? In that case, it should be collapsed into just one data point per country.

```{r load_fisheries_data}
fisheries <- read_csv("data/fisheries.csv")
```

Looking at the data, there are way to many countries to graph. Therefore, I will try seeing what the top countries are based on total fishery production:

```{r arrange_top}
fisheries %>%
  arrange(desc(total))
```

I think top 10 countries might be good. There might be political reasons to include other specific countries, but I don't have any of those, so I will stick with top 10. 

```{r filter_top_10}
fisheries_top_10 <- fisheries %>%
  slice_max(order_by = total, n = 10)
fisheries_top_10
```

Making a bar graph:
```{r plot_fish_bar_graph}
fisheries_top_10 %>%
  mutate(total = total / 100000) %>%
  ggplot(aes(x = reorder(country, -total), y = total, fill = country)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    total = "Total Fish Harvest by Top 10 Countries",
    x = "Countries",
    y = "Total Harvest (in 100 000s)"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

It could also be interesting to see the distribution of fish captures and harvested for each country in the bar graph. I will make that version of the plot too. 

```{r dist_fish_bar_graph}
# Make a new data set that is long instead of wide to make the capture and aquaculture variables one variable that can be stacked by
fisheries_top_10_long <- fisheries_top_10 %>%
  pivot_longer(cols = c(capture, aquaculture), names_to = "type", values_to = "tons_fish")

# Creating the stacked bar plot
fisheries_top_10_long %>%
  # Mutating to make the variable in 100 000s, so that the scale is more readable
  mutate(tons_fish = tons_fish / 100000) %>%
  ggplot(aes(x = reorder(country, -total), y = tons_fish, fill = type)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  labs(
    total = "Total Fish Harvest by Top 10 Countries",
    x = "Countries",
    y = "Total Harvest (in 100 000s)"
  ) +
  theme_minimal()
```


### Exercise 4

```{r load-data-Whickham}
data(Whickham)
```


#### Question 1

I think the data comes from an observational study, because they are looking at smoking and survival rate. It would hard to design an ethical study where some people will smoke regularly and others won't, so it is most like observational.


#### Question 2

There are `r nrow(Whickham)` observations in the data set. Each observation is a participant in the study.


#### Question 3
There are `r ncol(Whickham)` variables in the data set. These are `r colnames(Whickham)`. The object type of the columns are: `rsapply(Whickham, class)`.

```{r plot-outcome}
Whickham %>%
  ggplot(aes(x = outcome, fill = outcome)) +
  geom_bar() +
  labs(
    title = "Survival Rates in the Whickham data set",
    x = "Outcome",
    y = "Number of individuals"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

```{r plot-smoker}
Whickham %>%
  ggplot(aes(x = smoker, fill = smoker)) +
  geom_bar() +
  labs(
    title = "Skoming Rates in the Whickham data set",
    x = "Smoker status",
    y = "Number of individuals"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

```{r plot-age}
Whickham %>%
  ggplot(aes(x = age)) +
  geom_histogram(fill = "darkblue") +
  labs(
    title = "Age for Participants in the Whickham data set",
    x = "Age",
    y = "Number of individuals"
  ) +
  theme_minimal()
```

#### Question 4

I would expect there to be a relationship between smoking and health outcome, so that smokers are more likely to not be alive at the time of data collection than non-smokers.


#### Question 5

```{r plot-rel}
Whickham %>%
  count(smoker, outcome) %>%
  ggplot(aes(x = outcome, y = n, group = smoker, color = smoker)) +
  geom_line(linewidth = 1.5) +
  labs(
    title = "Survival Status based on Smoking Status",
    x = "Alive or dead",
    y = "Number of individuals"
  ) +
  theme_minimal()
```

There doesn't really seem to be any relationship between smoking and survival outcome as the lines are more or less parallel. This is not what I was expecting. 

```{r calc-prob}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(cond_prob = n /sum(n))

# Got some help from ChatGPT to figure out how to use the group_by and mutate to make the new variable without making a function. I apparently don't understand group_by as well as I could.
```

The probabilities of being alive or dead are also fairly similar based on smoking status. It is even slightly more likely for participants that were smokers to be alive, than for non-smokers.


### Question 6

Creating a new variable for binned ages:
```{r age-cat-var}
Whickham_df <- Whickham %>%
  mutate(age_cat = case_when(
    age <= 44 ~ "18-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 ~ "65+"
  ))
```


### Question 7

Recreating the previous plot faceted by age category:
```{r plot-by-age}
Whickham_df %>%
  count(smoker, age_cat, outcome) %>%
  ggplot(aes(x = outcome, y = n, group = smoker, color = smoker)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(vars(age_cat)) +
  labs(
    title = "Survival Status based on Smoking Status and Age Category",
    x = "Alive or dead",
    y = "Number of individuals"
  ) +
  theme_minimal()
```

Now, you can see that smoking makes it more likely that you will not survive in the young age category, from 18-44. At 45-64 years old, smoking status doesn't really matter. While at 65+, smoking makes it more likely that you will be alive compared to being a non-smoker.

Calculating conditional probabilities:
```{r cond-prob-age}
Whickham_df %>%
  count(smoker, age_cat, outcome) %>%
  group_by(smoker, age_cat) %>%
  mutate(cond_prob = n /sum(n))

```

Looking at this data, I would want to look further into the sample sizes per category. There are for example very few participants that are still alive in the smoking category, which could influence the results. In addition, I would also think that there are other factors (confounds) that could influence the relationships we see here. It might be that people who are older (65+) and have bad health, stop smoking, so that it looks like not smoking makes you more likely to die, when in reality, it might just be that people in bad health stop smoking to try to improve their health (and at the same time tend to die earlier because of their bad health). In this case, there might be a correlation, but not causation, between smoking and survival outcome, since there is a third variable, health, that effects both of our variables, making it look like they affect each other.