---
title: "Lab 03 - Nobel laureates"
author: "Linn Zapffe"
date: "28/1/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
nobel <- read_csv("data/nobel.csv")
```

## Exercises

### Exercise 1

The data frame has `r nrow(nobel)` observations and `r ncol(nobel)` columns. Each row is one winner of the Nobel price.

### Exercise 2

Making a new data frame called nobel_living, filtering after several criteria
```{r create-nobel_living}
nobel_living <- nobel %>%
    # Filter by if they have a country listed
  filter(!is.na(country)) %>%
    # Filter out organizations
  filter(gender != "org") %>%
    # Filter out individuals that are dead
  filter(is.na(died_date))
```

Trying to do the same with one filter statement:
```{r create-nobel_living-sec}
nobel %>%
    # Filter by several conditions in one argument
  filter(!is.na(country) & gender != "org" & is.na(died_date))
```

Both of the above codes worked.

The new data frame has `r nrow(nobel_living)` observations.

### Exercise 3

New variable to see whether the winners were in the US when winning the price
```{r mut_liv_US}
nobel_living <- nobel_living %>%
  mutate(
    country_us = if_else(country == "USA", "USA", "Other")
  )
```

Creating new variable for area of science
```{r mut_area}
nobel_living_science <- nobel_living %>%
  filter(category %in% c("Physics", "Medicine", "Chemistry", "Economics"))
```

```{r plot_box_area}
ggplot(nobel_living_science, mapping = aes(x = country_us)) +
  geom_bar() +
  facet_wrap(~category) +
  coord_flip()
```

Most of the winners were in the US when they won, but many of them were also outside of the US. I don't know whether looking at only category of price and where they won the price has anything to say about immigration. Need more information.

### Exercise 4

```{r var_born}
nobel_living_science <- nobel_living_science %>%
  mutate(
    born_country_us = if_else(born_country == "USA", "USA", "Other")
  )
```

Out of the winners, `r nobel_living_science %>% summarize(count = sum(born_country_us == "USA"))` were born in the US and `r nobel_living_science %>% summarize(count = sum(born_country_us == "Other"))` were born in other countries.

### Exercise 5

```{r plot_area_and_born}
ggplot(nobel_living_science, mapping = aes(x = country_us, fill = born_country_us)) +
  geom_bar() +
  facet_wrap(~category) +
  coord_flip()
```

Most of the winners in the US were born in the US, but there is a pretty significant portion of winners that were born other places. I would say that Buzzfeed has a point about immigrants being good for the research environment in the US.

### Exercise 6

```{r freq_plot_country}
nobel_living_science %>% 
  filter(country != "USA" & born_country_us == "Other") %>%
  count(born_country) %>%
  arrange(desc(n))
```

The most common country for immigrants was Japan, with 14 winners.