---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Linn Zapffe"
date: "29/1/2025"
output: github_document
---

### Load packages and data

```{r install_dsbox}
# Ran this code and then commented it out so that it isn't re-run everytime this document is knitted
#install.packages("devtools")
#devtools::install_github("rstudio-education/dsbox")
```


```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

The dimensions of the Denny's data set are `r nrow(dennys)` rows and `r ncol(dennys)` columns. Each row is a Denny's location and the 6 variables are: `r names(dennys)`.


### Exercise 2

The dimensions of the La Quinta data set are `r nrow(laquinta)` rows and `r ncol(laquinta)` columns. Each row is a La Quinta location and the 6 variables are: `r names(laquinta)`. These variables are the same as for the Denny's locations data set.

### Exercise 3

Whether there are any Denny's locations outside of the US depends on the definition. Isn't Texas its own country? If not, then there are no Denny's outside of the US.

La Quinta has locations all over the world, even though most of them are in the US. The countries are Canada, Mexico, China, New Zealand, Turkey, United Arab Emirates, Chile, Columbia, and Equador.

### Exercise 4

I would assume the state would be missing if the location is outside of the US, so I would filter on state being missing.

### Exercise 5

```{r filter_dennys_not_US}
dennys %>%
  filter(!(state %in% states$abbreviation))
```

There are no Denny's locations outside of the US.

### Exercise 6

```{r dennys_country_var}
dennys <- dennys %>%
  mutate(country = "United States")
```


 
### Exercise 7

```{r filter_laquinta_not_US}
laquinta %>%
  filter(!(state %in% states$abbreviation))
```

The countries were listed already in question 3, see for reference.

### Exercise 8

```{r add_country_var_laquinta}
laquinta <- laquinta %>%
  mutate(country = case_when(
    state %in% state.abb ~ "United States",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT" ~ "Colombia",
    state %in% c("AG", "QR", "CH", "NL", "VE", "PU", "SL") ~ "Mexico",
    state == "FM" ~ "Honduras"
  ))
```

This countries I found by Googling did not seem to be the same as from the La Quinta webpage. There was no Honduras on the webpage and there are a lot of countries on the webpage missing in the data set. My best guess is that the data set is older than the webpage and not updated with all  the current locations.

```{r filter_US}
laquinta <- laquinta %>%
  filter(country == "United States")
```

### Exercise 9

```{r most_locations_states_dennys}
dennys %>%
  count(state) %>%
  arrange(desc(n))
```

There are the most Denny's locations in California.

```{r most_locations_states_laquinta}
laquinta %>%
  count(state) %>%
  arrange(desc(n))
```

There are the most La Quinta locations in Texas.

It is a little surprising that they don't have the highest number in the same state, since they are (argued) to always occur together. Other than that, I don't know enough about the different states or stereotypes of the different states to evaluate the specific state further.

### Exercise 10

```{r loc_per_mile_dennys}
dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>% 
  mutate(loc_per_mile = area / 1000 / n) %>% 
  arrange(loc_per_mile)
```

We used inner join here as it found each row that matched (the US states) and kept the variables from the two data sets from those rows.

There are the most locations of Denny's per thousand square miles in the District of Columbia. I didn't even know that that was a state.


```{r loc_per_mile_laquinta}
laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>% 
  mutate(loc_per_mile = area / 1000 / n) %>% 
  arrange(loc_per_mile)
```

There are the most of La Quinta locations per thousand square miles in Rhode Island.


### Exercise 11

```{r type_var}
dennys <- dennys %>%
  mutate(establishment = "Denny's")
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")
```

```{r added}
dn_lq <- bind_rows(dennys, laquinta)
```

```{r scatterplot_map}
ggplot(dn_lq, mapping = aes(
  x = longitude,
  y = latitude,
  color = establishment
)) +
  geom_point()
```

```{r scatter_plot_NC}
dn_lq %>% 
  filter(state == "NC") %>% 
  ggplot(dn_lq, mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
    geom_point(alpha = 0.25) +
  labs(
    title = "Geographical Locations of Denny's and La Quintas in North Carolina",
    color = "Establishment",
    x = "Longitude",
    y = "Latitude"
  )
```

The joke doesn't really seem to hold based on the locations in North Carolina, as there are only a few locations at the same place.


### Exercise 12

```{r scatter_plot_TX}
dn_lq %>% 
  filter(state == "TX") %>% 
  ggplot(dn_lq, mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
    geom_point(alpha = 0.25) +
  labs(
    title = "Geographical Locations of Denny's and La Quintas in Texas",
    color = "Establishment",
    x = "Longitude",
    y = "Latitude"
  )
```

The claim seems more supported in Texas, as there are clusters with these locations, where several of them seem to overlap (i.e. being in the same location).